Learning how to act: making good decisions with machine learning

### 2017-09-07 thesis/thesis_map.tex
adds minor fixes up to start of 2211

The goal of this thesis to connect and unify these key approaches. I introduce causal bandit problems: a framework that combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcement learning problems that are simple enough to admit formal analysis. I show that knowledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be expoited by existing algorithms. I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and strictly better than algorithms that do not use the additional causal information.

@@

There has traditionally been a large gap between researchers in machine learning who focus on prediction, using largely non-interpretable models and researchers in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use highly theory-driven models. However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation models. In the next section we emphasise the subtley that can exist in determining if a problem is causal by examining some typical examples.

@@

@@

Demand prediction seems like a relatively straightforward prediction problem. These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product. It seems unlikely that using the model to ensure stock is available will itself change demand. However, depending on the way demand is measured, there is a potential data censoring issue. If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero. Changing availability does then change demand.

Playing Go (and other games) is a case with some subtleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem. There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win, it must identify higher level features of the board state that are associated with winning.}. Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of selecting action $a$ given board state $s$. This allows the agent to estimate the likelyhood of winning from a given starting state by simulating (many times) the remainder of the game, drawing actions from $\P{a|s}$ for both players. Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a combination of these approaches \citep{silver2016mastering}. The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.

@@

Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predictive) inference. A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes. The second requirement is particularly difficult to satisfy when an algorithm is making important decisions effecting individual people. Think about problems like credit scoring and parole decisions. There are strong ethical grounds for demanding transparency, but if the goals of society and the individuals are not perfectly aligned and there is any possibility that people can manipulate features independently of the outcome, there will be a conflict between model accuracy and transparency. It is rare to build a model without any intent to make some kind of decision on the basis of its results. Thus, I argue we should assume a causal model is required until we can justify otherwise.

@@

The approach of learning the outcome of an action by taking it has plays a key role in advancing our knowledge of the world. However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen. Estimating the affect of an action from such observational data sets is the problem addressed by observational causal inference. Observational causal inference can be viewed as a form of transfer learning. The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the original system. This is impossible without some assumptions about how the the original system and the system after intervention are related to one-another. The key to observational inference is to model how actions change the state of the world in such a way that we can map information collected in one setting to another.

Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials. Bandit algorithms deal with the sequential nature of the decision making process, causal inference with the problem randomisation is not always feasible, affordable or ethical. The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined.

@@

I view causal problems as one of the greatest arising challenges for machine learning. They incorporate a large set of problems, of huge practical significance, that require us to go beyond pattern recognition, but are well short of general artificial intelligence. For the major advances in representation and pattern recognition developed within machine learning to be effectively applied in many areas of medicine, economics, social sciences and industry we need to understand how to leverage our improved approaches to prediction to tackle causal problems.

\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints. My major contribution is a framework which unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty. My other contributions are primarily connective rather than technical. I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias. I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data and covariate shift. I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.

@@

Chapter \ref{chap:interventionalist_view} deals with interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems. In \S\ref{sec:multi-armed-bandits} I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}). I clarify the causal structure of (stochastic) contextual bandit problems in \S\ref{sec:causal-structure-contextual-bandit}. In \S\ref{sec:off-policy-evaluation} I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of of causal effect estimation from observational data. Finally, in section \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature bandit regret.

In chapter \ref{chap:causal-bandit} I introduce causal bandit problems: a framework which unifies causal graphical models with multi-armed bandit problems. Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model. I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of bandit problem depending what information is observable and when.

In \S\ref{sec:causal-bandit-post-action-feedback} I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected. I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms. In \S\ref{sec:simple-regret} I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}. I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning. In \S\ref{sec:simple-regret-general} I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}. Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems. Finally in \S\ref{sec:causal_bandit_discussion} I discuss extensions and potential future work.

@@

The goal of causal inference is to learn the effect of taking an action. We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world. We are generating and storing data on almost every aspect of our lives at an unprecedented rate. As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase. However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question. An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited. This makes it critical that we develop effective methods that us to predict the outcome of an intervention in some system from observing it. This is the problem of observational causal inference.

@@

Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings. They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action verses what would happen if we took another. Counterfactual assertions can be expressed very naturally in human languages and are quite prevalent in  every day conversations;\quotes{if I had worked harder I would have got better grades}, \quotes{she would have been much sicker if she hadn't taken antibiotics}. Structural equation models have been developed and applied primarily within economics and related disciplines and can be seen as an attempt to capture key aspects of the behaviour of people with mathematics. Questions around designing policies or interventions play a central role in economics. Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised) framework and associated set of methods for estimating causal effects. Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks. They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of conditions under causal effects can be inferred from observational data.

An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}. However, the literature on causal inference techniques remains spit between the different frameworks. Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scores, are described using the counterfactual framework. Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions are often based around structural equation models. This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.

In the next sections we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they allow us to solve, the assumptions they rely on and how they differ. We hope that by describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them. However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped by uninterested readers. In order to demonstrate the notation and formalisms each framework provides we will use them to describe the following simple examples.

@@

A causal Bayesian network is a Bayesian network in which a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$. This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change. We need some notation to describe interventions and represent distributions over variables in the network after an intervention. In this thesis I use the do operator introduced by Pearl \citep{Pearl2000}.

\vspace{0.5cm}
\begin{definition}{The do-notation}
\begin{itemize}
\item $do(X=x)$ denotes an intervention that sets the random variable(s) $X$ to $x$.
\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$. This notation is somewhat overloaded. It may be used to represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed. For example it could represent 
\begin{itemize}
\item the probability $\P{Y=1|do(X=x)}$ as a function of $x$,
\item the probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$,
\item the probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$,
\item a family of density/mass function for $Y$ parameterised by $x$.
\end{itemize}
Where the distinction is important and not clear from context we will use one of the more specific forms above. 
\end{itemize}
\end{definition}

@@

Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$? The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient. Since $X$ is assigned via deliberate randomisation we know that it is not affected by any latent variables. The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}. This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$. In this example, the interventional distribution is equivalent to the observational one.

@@

In example \ref{exm:adjusting} we are interested in $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool. We could compare it to outlawing preschool $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$. It seems reasonable to assume that preschool attendance affects the likelihood of high school graduation \footnote{The effect does not have to be homogeneous, it may depend non-linearly on characteristics of the child, family and school.} and that parental socio-economic status would affect \emph{both} the likelihood of preschool attendance and high school graduation. If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal Bayesian network in figure \ref{fig:causal_adjust}. In this case, the interventional distribution is not equivalent to the observational one. If parents with high socio-economic status are more likely to send their children to preschool and these children are more likely to graduate high school regardless, comparing the graduation rates of those who attended preschool with those who did not will overstate the benefit of preschool. To obtain the interventional distribution we have to estimate the impact of preschool on high school graduation for each socio-economic level separately and then weight the results by the proportion of the population in that group,

@@

A causal Bayesian network represents much more information than a Bayesian network with identical structure. A causal network encodes all possible interventions that could be specified with the do-notation. For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters. The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions. Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional 30 parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters. After each two variable intervention, we have a distribution over one variables which requires one parameter. This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.

@@

\subsubsection{Limitations of causal Bayesian networks}
A number of criticisms have been levelled at this approach to modelling causality. One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}. However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions. For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter. But, what if some people failed to take the prescribed treatment? We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}. Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.

@@

Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables. Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology. We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on. However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency. Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference. Such problems have mostly been studied within the dynamical systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable. \citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.

@@

There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised. In particular, statements involving joint distributions over counterfactual variables may not be able to be validated empirically \citet{Dawid2000}. One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation. Rather than talking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$. This is certainly convenient, if rather imprecise. However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrants careful attention. It is important to be clear what assumptions are being made and whether or not they could be validated (at least in theory).

@@

The interventional distribution is the same as the observational one. The same argument applies to the intervention $do(X=0)$ and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.

@@

The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals. When discussing the counterfactual model, we actually made the slightly weaker assumption:

>>>>

The goal of this thesis to connect and unify these key approaches. I introduce causal bandit problems: a framework that combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcement learning problems that are simple enough to admit formal analysis. I show that knowledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be exploited by existing algorithms. I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and strictly better than algorithms that do not use the additional causal information.

@@

There has traditionally been a large gap between researchers in machine learning who focus on prediction, using largely non-interpretable models and researchers in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use highly theory-driven models. However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation models. In the next section we emphasise the subtly that can exist in determining if a problem is causal by examining some typical examples.

@@

Predicting recidivism among paroled prisoners or customer churn also fit within the class of problems where the goal is to identify a group for which a problem will occur in order to target treatment (additional support and monitoring for people on parole, loyalty rewards to improve customer retention, hospitalisation for the severely ill). Predictive models can be applied to such problems where the most effective treatment is known for a given target group, and where deciding who to treat on the basis of the model predictions will not change the relationship between the features and outcome.

@@

Demand prediction seems like a relatively straightforward prediction problem. Models use features such as location, pricing, marketing, time of year and weather, to forecast the demand for a product. It seems unlikely that using the model to ensure stock is available will itself change demand. However, depending on the way demand is measured, there is a potential data censoring issue. If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero. Changing availability does then change demand.

Playing Go (and other games) is a case with some subtleties. At every turn, the AI agent has a number of actions available. The state of the board following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem. There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win. It must identify higher level features of the board state that are associated with winning.}. Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of a player selecting action $a$, given board state $s$. This allows the agent to estimate the likelihood of winning from a given starting state by simulating many times the remainder of the game, drawing actions from $\P{a|s}$ for both players. Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a combination of these approaches \citep{silver2016mastering}. The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelihood an opponent will play a given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.

@@

Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predictive) inference. A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes. The second requirement is particularly difficult to satisfy when an algorithm is making important decisions affecting individual people. Think about problems like credit scoring and parole decisions. There are strong ethical grounds for demanding transparency, but if the goals of society and the individuals are not perfectly aligned and there is any possibility that people can manipulate features independently of the outcome, there will be a conflict between model accuracy and transparency. It is rare to build a model without any intent to make some kind of decision based on its results. Thus, I argue we should assume a causal model is required until we can justify otherwise.

@@

The approach of learning the outcome of an action by taking it plays a key role in advancing our knowledge of the world. However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen. Estimating the effect of an action from such observational data sets is the problem addressed by observational causal inference. Observational causal inference can be viewed as a form of transfer learning. The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the original system. This is impossible without some assumptions about how the the original system and the system after intervention are related to one-another. The key to observational inference is to model how actions change the state of the world in such a way that we can map information collected in one setting to another.

Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials. Bandit algorithms deal with the sequential nature of the decision making process, and causal inference with problems where randomisation is not feasible, affordable or ethical. The similarities between the problems addressed by these techniques raises the question of whether there are problems best addressed by a combination of these approaches, and if so, how they can be combined.

@@

I view causal problems as one of the greatest current challenges for machine learning. They incorporate a large set of problems of huge practical significance, that require us to go beyond pattern recognition, but are well short of general artificial intelligence. For the major advances in representation and pattern recognition developed within machine learning to be effectively applied in many areas of medicine, economics, social sciences and industry, we need to understand how to leverage our improved approaches to prediction to tackle causal problems.

\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints. My major contribution is a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty. My other contributions are primarily connective rather than technical. I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias. I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data, and covariate shift. I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.

@@

Chapter \ref{chap:interventionalist_view} deals with interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems. In \S\ref{sec:multi-armed-bandits}, I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}). I clarify the causal structure of (stochastic) contextual bandit problems in \S\ref{sec:causal-structure-contextual-bandit}. In \S\ref{sec:off-policy-evaluation}, I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of causal effect estimation from observational data. Finally, in section \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature bandit regret.

In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems: a framework that unifies causal graphical models with multi-armed bandit problems. Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model. I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of bandit problem depending what information is observable and when.

In \S\ref{sec:causal-bandit-post-action-feedback}, I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected. I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms. In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}. I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning. In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}. Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems. Finally in \S\ref{sec:causal_bandit_discussion}, I discuss extensions and potential future work.

@@

The goal of causal inference is to learn the effect of taking an action. We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world. We are generating and storing data on almost every aspect of our lives at an unprecedented rate. As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase. However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question. An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited. This makes it critical that we develop effective methods that enable us to predict the outcome of an intervention in some system from observing it. This is the problem of observational causal inference.

@@

Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings. They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action verses what would happen if we took another. Counterfactual assertions can be expressed very naturally in human languages and are  prevalent in  everyday conversations; \quotes{if I had worked harder I would have got better grades} and \quotes{she would have been much sicker if she hadn't taken antibiotics}. Structural equation models have been developed and applied primarily within economics and related disciplines. They can be seen as an attempt to capture key aspects of the people's behaviour with mathematics. Questions around designing policies or interventions play a central role in economics. Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised \citep{Pearl2000}) framework and associated set of methods for estimating causal effects. Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks. They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of the conditions under which causal effects can be inferred from observational data.

An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}. However, the literature on causal inference techniques remains spit between the different frameworks. Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scoring, are described using the counterfactual framework. Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions, are often based around structural equation models. This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.

In the next sections, we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they allow us to solve, the assumptions they rely on and how they differ. By describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them. However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped by uninterested readers. In order to demonstrate the notation and formalisms each framework provides, we will use them to describe the following simple examples.

@@

A causal Bayesian network is a Bayesian network in which a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$. This means an intervention to change the value of $X_{i}$ can be expected to affect $X_{j}$, but interventions on $X_{j}$ will not affect $X_{i}$. We need some notation to describe interventions and represent distributions over variables in the network after an intervention. In this thesis, I use the do operator introduced by Pearl \citep{Pearl2000}.

\vspace{0.5cm}
\begin{definition}{The do-notation}
\begin{itemize}
\item $do(X=x)$ denotes an intervention that sets the random variable(s) $X$ to $x$.
\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$. This notation is somewhat overloaded. It may be used to represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed. For example, it could represent 
\begin{itemize}
\item the probability $\P{Y=1|do(X=x)}$ as a function of $x$,
\item the probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$,
\item the probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$,
\item a family of density/mass function for $Y$ parameterised by $x$.
\end{itemize}
Where the distinction is important and not clear from context we will use one of the more specific forms above. 
\end{itemize}
\end{definition}

@@

Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$? The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient. Since $X$ is assigned via deliberate randomisation, it cannot be affected by any latent variables. The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}. This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$. In this example, the interventional distribution is equivalent to the observational one.

@@

In example \ref{exm:adjusting} we are interested in $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool. We could compare it to outlawing preschool $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$. It seems reasonable to assume that preschool attendance affects the likelihood of high school graduation \footnote{The effect does not have to be homogeneous, it may depend non-linearly on characteristics of the child, family and school.} and that parental socio-economic status would affect \emph{both} the likelihood of preschool attendance and high school graduation. If we assume that socio-economic status is the only such variable (nothing else affects both attendance \emph{and} graduation), we can represent this problem with the causal Bayesian network in figure \ref{fig:causal_adjust}. In this case, the interventional distribution is not equivalent to the observational one. If parents with high socio-economic status are more likely to send their children to preschool and these children are more likely to graduate high school regardless, comparing the graduation rates of those who attended preschool with those who did not will overstate the benefit of preschool. To obtain the interventional distribution we have to estimate the impact of preschool on high school graduation for each socio-economic level separately and then weight the results by the proportion of the population in that group,

@@

A causal Bayesian network represents much more information than a Bayesian network with identical structure. A causal network encodes all possible interventions that could be specified with the do-notation. For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by seven parameters. The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions. Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional thirty parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters. After each two variable intervention, we have a distribution over one variables which requires one parameter. This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.

@@

\subsubsection{Limitations of causal Bayesian networks}
A number of criticisms have been levelled at this approach to modelling causality. One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}. However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions. For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter. But, what if some people failed to take the prescribed treatment? We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive (figure \ref{fig:randomized_imperfect_compliance}). Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.

@@

Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables. Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology. We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on. However, if our measurements of these variables must be aggregated over time periods that are longer than the scale at which these interactions occur, the result is a cyclical dependency. Even were we able to measure on shorter timescales, there might then not be sufficient data on each variable for inference. Such problems have mostly been studied within the dynamical systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable. \citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and that can be seen as Bayesian causal networks with additional functional assumptions.

@@

There are objections to counterfactuals arising from the way they describe alternate universes that were never realised. In particular, statements involving joint distributions over counterfactual variables may not be able to be validated empirically \citet{Dawid2000}. One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation. Rather than talking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$. This is certainly convenient, if rather imprecise. However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrants careful attention. It is important to be clear what assumptions are being made and whether or not they could be validated (at least in theory).

@@

The interventional distribution is the same as the observational one. The same argument applies to the intervention $do(X=0)$, and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.

@@

The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals. When discussing the counterfactual model, we made the slightly weaker assumption:
